Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-02-12 Paper.Rmd
edits to Q3 sections

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs; and likewise for SG and SJ copulas. Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring - apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

@@

We addressed Q3 using both models and data. The initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should depend on the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Using similar reasoning, left-tail dependence between local variables
$E_i$ and $E_j$ will tend to decreases skewness of the spatial mean. Skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series. 
The spatial-mean time series may be the quantity of principal importance for 
cases such as pest or resource abundance or CO$_2$ or methane flux.

We tested these ideas using the focal multivariate datasets of our earlier analyses:
green spruce aphid abundance data, leaf-curling plum aphid first flight data, \emph{Ceratium
furca} abundance data, and methane flux data (see Data and Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean, and then the skewness (through time) of that
mean. Then, for each dataset separately, we compared the value obtained 
to a distribution of values 
of the same quantity calculated for each of 10000 surrogate datasets that were 
produced from the empirical data by randomizing
it in a special way to have the copula structure of a multivariate-normal distribution
(Appendix \ref{SM-Surrog_multi}). The surrogate/randomized datasets had exactly the same
marginal distributions (i.e., distributions of values for each sampling location)
as the original data. 
They also had very similar Kendall or Spearman correlations (the algorithm had two versions,
one for preserving each correlation coefficient) between pairs of sampling locations
as the original data. But the surrogates had multivariate normal copula structure. 
Our comparisons therefore tested the null hypothesis that the skewness of the spatial 
mean took values on the empirical no different from what one would expect if the copula
structure of the data were multivariate normal, but the data were otherwise statistically
unchanged. Significant differences indicate that copula structure contributes to the skewness
of the spatial mean, i.e., to its "spikiness."

<!--***DAN: stopped here 2019 02 08 1106am central-->

This section is dedicated to analysis of question \hyperref[Intro]{\textbf{(Q3)}} from \nameref{Intro}. We wanted to examine whether the asymmetric tail dependence had any **(i)** effects on skewness of spatially averaged time series data. We predicted that because of overall distinct tail dependence of the multivariate copula structure, the skewness of the spatially-averaged time-series data, should change significantly from the individual skewness of time-series data. We tested null hypothesis $(H_{0})$ which presumed that multivariate copula actually came from normal copula structure with no tail dependence and thus should have the same skewness as of normal copula.
We generated $10000$ number of surrogate normal copula from a given multivariate copula keeping the marginal distribution and the correlation (Spearman or Kendall) same and calculated the skewness of spatially averaged surrogate time series data, as we did with the actual multivariate data set in the next section (For code see [skewness_multivar_dataset.R](https://github.com/sghosh89/BIVAN/blob/master/skewness_multivar_dataset.R)). We made a histogram of surrogate skewness and found if actual skewness was within the $95\%$ confidence interval or not. If the $p$-value is less than $0.05$, then we reject the null hypothesis in support of significant tail dependence (and henceforth significantly different skewness than Normal copula).

We also showed **(ii)** asymmetric copula structure affected slope of spatial Taylor's Law differently than as expected from symmetric Normal copula (For details see Appendix \ref{SM-Consequence_TL_simulation}). For example, our model simulation showed lower tail dependent Clayton copula had much higher slope (spatial-Taylor's law) than symmetric Normal copula, which in turn had higher slope than upper tail dependent Survival Clayton family (Fig. \ref{SM-fig_spTL}(I)). In the following section we will show application of these ideas with real data sets.
<!-- Extinction risk results are not shown currently, it is introduced as a future idea in Discussion-----------
Finally, we studied **(iii)** the effects of environmental noise, having a non-Normal copula, on extinction risk (calculated after 25 years) with two metapopulation model, namely, Lewton-Cohen model and Ricker model (For details see Appendix \ref{SM-Consequence_ext_risk}). We considered patches were arranged in a linear chain fashion and the population may disperse to its nearerst two neighbors (local dispersion) or to any patches (global dispersion). For both model with both dispersion, increasing number of habitat patches and increasing carrying capacity, separately, had decreased extinction risk of the populations which was expected (Fig. \ref{SM-fig_risk_numlocs_effect_local_disp} - Fig. \ref{SM-fig_risk_K_effect}). For Lewton-Cohen model, we found greater extinction risk if populations were affected by left tail dependence environmental noise copula than right tail dependent one. For Ricker model, we got the similar observation for smaller values of $r$, the intrinsic growth rate, but for higher values of $r$, there was a turn-over in extinction risks, and right tail dependent noise copula caused population go extinct faster. -->

\begin{figure}[!h]
\textbf{ \hspace{6 cm} (A) \hspace{6 cm} (B)} \\
\vspace{-0.1 cm}
\begin{center}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_aphid_count/skewness_aphid_count_sp_10_spearman.pdf}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_aphid_ff/skewness_aphid_ff_sp_11_spearman.pdf}\\
\end{center}
\vspace{0.2 cm}
\textbf{ \hspace{6 cm} (C) \hspace{6 cm} (D)} \\
\vspace{-0.1 cm}
\begin{center}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_plankton_north_sea/skewness_plankton_north_sea_sp_16_spearman.pdf}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_methane/skewness_methane_spearman.pdf}
\caption{Skewness of spatially averaged $(A)$ green-spruce aphid abundance, $(B)$ leaf-curling plum aphid first flight date, $(C)$ Ceratium furca abundance and $(D)$ methane-flux data, were indicated by red dot in each plot and were compared with respect to skewness of spatially averaged 10000 Normal surrogates preserving Spearman correlation). \label{fig_skewness_spearman}}
\end{center}
\end{figure}

## Results for \hyperref[Intro]{\textbf{(Q3)}}: Change of skewness

>>>>

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs; and likewise for SG and SJ copulas. Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

@@

We addressed Q3 using both models and data. An initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should be influenced by the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Similarly, left-tail dependence between local variables
will tend to decreases skewness of the spatial mean. Skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series. 
The spatial-mean time series may be the quantity of principal importance for 
cases such as pest or resource abundance or CO$_2$ or methane flux.

We tested these ideas using the focal multivariate datasets of our earlier analyses:
green spruce aphid abundance data, leaf-curling plum aphid first flight data, \emph{Ceratium
furca} abundance data, and methane flux data (see Data and Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean, and then the skewness (through time) of that
mean. Then, for each dataset separately, we compared the value obtained 
to a distribution of values 
of the same quantity calculated for each of 10000 surrogate datasets that were 
produced from the empirical data by randomizing
it in a special way to have the copula structure of a multivariate-normal distribution
(Appendix \ref{SM-Surrog_multi}). The surrogate/randomized datasets had exactly the same
marginal distributions (i.e., distributions of values for each sampling location)
as the original data. 
They also had very similar Spearman correlations between pairs of sampling locations
as the original data. But the surrogates had multivariate normal copula structure. 
Our comparisons therefore tested the null hypothesis that the skewness of the spatial 
mean took values on the empirical data no different from what one would expect if the copula
structure of the data were multivariate normal, but the data were otherwise statistically
unchanged. Significant differences indicate that copula structure contributed 
to the skewness of the spatial mean, i.e., to its "spikiness."

For green spruce aphid abundance data, *Ceratium furca* abundance data, and methane-flux
data, because these datasets exhibited lower-tail dependence 
(Table \ref{tab_multivar_summary}), we compared empirical and surrogate skewness values
via a one-tailed test in the left tail: the $p$-value result was simply the fraction
of surrogate skewnesses less than the skewness computed from the empirical data. For
leaf-curling plum aphid first flight data, because this dataset exhibited upper-tail 
dependence (Table \ref{tab_multivar_summary}), we did the analogous one-tail test
in the right tail.

<!--Reminder that Table 2 says:
Green spruce - lower tail
Leaf-curling - upper tail
Cerratium - lower tail
methane - marginal lower tail-->

<!--***DAN: stopped here in this section, 2019 02 12, 633am-->

We also showed **(ii)** asymmetric copula structure affected slope of spatial Taylor's Law differently than as expected from symmetric Normal copula (For details see Appendix \ref{SM-Consequence_TL_simulation}). For example, our model simulation showed lower tail dependent Clayton copula had much higher slope (spatial-Taylor's law) than symmetric Normal copula, which in turn had higher slope than upper tail dependent Survival Clayton family (Fig. \ref{SM-fig_spTL}(I)). In the following section we will show application of these ideas with real data sets.

<!-- Extinction risk results are not shown currently, it is introduced as a future idea in Discussion-----------
Finally, we studied **(iii)** the effects of environmental noise, having a non-Normal copula, on extinction risk (calculated after 25 years) with two metapopulation model, namely, Lewton-Cohen model and Ricker model (For details see Appendix \ref{SM-Consequence_ext_risk}). We considered patches were arranged in a linear chain fashion and the population may disperse to its nearerst two neighbors (local dispersion) or to any patches (global dispersion). For both model with both dispersion, increasing number of habitat patches and increasing carrying capacity, separately, had decreased extinction risk of the populations which was expected (Fig. \ref{SM-fig_risk_numlocs_effect_local_disp} - Fig. \ref{SM-fig_risk_K_effect}). For Lewton-Cohen model, we found greater extinction risk if populations were affected by left tail dependence environmental noise copula than right tail dependent one. For Ricker model, we got the similar observation for smaller values of $r$, the intrinsic growth rate, but for higher values of $r$, there was a turn-over in extinction risks, and right tail dependent noise copula caused population go extinct faster. -->

## Results for Q3: Tail dependence influences skewness of spatial average, Taylor's law, and extinction risk

\begin{figure}[!h]
\textbf{ \hspace{6 cm} (A) \hspace{6 cm} (B)} \\
\vspace{-0.1 cm}
\begin{center}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_aphid_count/skewness_aphid_count_sp_10_spearman.pdf}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_aphid_ff/skewness_aphid_ff_sp_11_spearman.pdf}\\
\end{center}
\vspace{0.2 cm}
\textbf{ \hspace{6 cm} (C) \hspace{6 cm} (D)} \\
\vspace{-0.1 cm}
\begin{center}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_plankton_north_sea/skewness_plankton_north_sea_sp_16_spearman.pdf}
\includegraphics[width=5.5 cm]{./Results/skewness_results/skewness_methane/skewness_methane_spearman.pdf}
\caption{Skewness of spatially averaged green spruce aphid abundance $(A)$, leaf-curling plum 
aphid first flight dates $(B)$, \emph{Ceratium furca} abundance $(C)$, and methane-flux
$(D)$ time series compared to an multivariate 
N-copula null hypothesis. Red dots are empirical 
skewnesses; see text 
and Appendix \ref{SM-Surrog_multi} for details of the null hypothesis. Results show a 
tendency for skewness of the spatial average to be affected as hypothesized 
by tail dependence. \label{fig_skewness_spearman}}
\end{center}
\end{figure}
<!--***DAN: Shya, please replace the x axis label by "Skewness" and the y-axis label
by "Frequency"-->

<!--***DAN: stopped here in this section, 2019 02 12, 633am-->

### 2019-02-12 SupportingInformation.Rmd
edits to Q3 sections

```{r skewness_aphidcount_allsp_plot, echo=F, results="hide", warning=F, cache=T, cache.extra=list(seed,data_aphid_count,mtime("good_loclist.R"),mtime("skewness_multivar_dataset.R"),mtime("ncsurrog.R"),mtime("SkewnessAnd3CentMom.R"))}

@@

```{r skewness_aphidff_allsp_plot, echo=F, results="hide", warning=F, cache=T, cache.extra=list(seed,data_aphid_ff,mtime("good_loclist.R"),mtime("skewness_multivar_dataset.R"),mtime("ncsurrog.R"),mtime("SkewnessAnd3CentMom.R"))}

@@

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Spearman (or Kendall) 
correlation of $x_i(t)$ and $x_j(t)$.
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Spearman (or Kendall) correlations equal to the $\tau_{ij}$. This is possible
using the `iRho` function (or the `iTau function) of the `VineCopula` package
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the the multivariate normal distribution of the previous step.
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k$th smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is the surrogate dataset.

\noindent Our code for this algorithm is `ncsurrog` in the `BIVAN` repository 
(\url{https://github.com/sghosh89/BIVAN}). Because, for each $i$, the final time series 
$b_i(t)$ is a permuted version of $x_i(t)$ (and with `NA`s for the same $t$), 
the marginal distributions of the surrogate dataset (i.e., the distributions of 
values for each sampling location, $i$) are exactly the same as for the original data.
Because ranks are preserved in passing from $a_i(t)$ to $b_i(t)$, pairwise Spearman 
(or Kendall) 
correlations for the $b_i(t)$ will be the same as those of the $a_i(t)$. 
Discrepancies between the $\tau_{ij}$ and the pairwise Spearman (or Kendall) 
correlations of the
$b_i(t)$ will therefore come about only from the presence of NAs and from sampling variation. 
These discrepancies should be minor if the 
numbers of NAs are not too large and $T$ is not too small. The Kendall version of the
algorithm throws an error if there are ties in any of the time series. The algorithm
also throws an error in the event that
the matrix generated in the second step is not positive semidefinite.

>>>>

```{r skewness_aphidcount_allsp_plot, echo=F, results="hide", cache=T, cache.extra=list(seed,data_aphid_count,mtime("good_loclist.R"),mtime("skewness_multivar_dataset.R"),mtime("ncsurrog.R"),mtime("SkewnessAnd3CentMom.R"))}

@@

```{r skewness_aphidff_allsp_plot, echo=F, results="hide", cache=T, cache.extra=list(seed,data_aphid_ff,mtime("good_loclist.R"),mtime("skewness_multivar_dataset.R"),mtime("ncsurrog.R"),mtime("SkewnessAnd3CentMom.R"))}

@@

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Spearman (or Kendall) 
correlation of $x_i(t)$ and $x_j(t)$.
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Spearman (or Kendall) correlations equal to the $\tau_{ij}$. This is possible
using the `iRho` function (or the `iTau function) of the `VineCopula` package
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the the multivariate normal distribution of the previous step.
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k$th smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is a surrogate dataset.

\noindent Our code for this algorithm is `ncsurrog` in the `BIVAN` repository 
(\url{https://github.com/sghosh89/BIVAN}).
The code for the Kendall version of the
algorithm throws an error if there are ties in any of the time series. This is
because the Kendall correlation in the very commonly used `cor` function in R 
is the Kendall tau-b coefficient in the event of ties, and index which 
differs in its definition from the commonly used tau-a that works when there are
no ties; and `tau` and `iTau` are intended for continuous copulas, which cannot 
produce data with ties. Our code also throws an error in the event that
the matrix generated in the second step is not positive semidefinite. Because
ties are common in our data, we only used the Spearman case in analyses using `ncsurrog`,
though, for expansibility, the `ncsurrog` code itself is written to handle 
the Kendall case as well when there are no ties.

Because, for each $i$, the final time series 
$b_i(t)$ is a permuted version of $x_i(t)$ (and with `NA`s for the same $t$), 
the marginal distributions of the surrogate dataset (i.e., the distributions of 
values for each sampling location, $i$) are exactly the same as for the original data.
Because ranks are preserved in passing from $a_i(t)$ to $b_i(t)$, pairwise Spearman 
(or Kendall) 
correlations for the $b_i(t)$ will be the same as those of the $a_i(t)$. 
Discrepancies between the $\tau_{ij}$ and the pairwise Spearman (or Kendall) 
correlations of the
$b_i(t)$ will therefore come about only from the presence of NAs and from sampling variation. 
These discrepancies should be minor if the 
numbers of NAs are not too large and $T$ is not too small.
